Project Log -- Severe Weather Impact Analysis (RDA Project 2)
=============================================================

Getting and loading the online data (cached)
--------------------------------------------

```{r Get and Load Online Data, cache = TRUE }
qtDload = TRUE
download.file( 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2',
               destfile = 'StormData.csv.bz2', method='curl', quiet = qtDload )
download.file( 'https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf',
               destfile = 'pd01016005curr.pdf', method='curl', quiet = qtDload )
download.file( 'https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf',
               destfile = 'Storm_Events_FAQ.pdf', method='curl', quiet = qtDload )
# install.packages("R.utils")
library('R.utils')
bunzip2( 'StormData.csv.bz2', remove = FALSE, overwrite = TRUE )
s <- read.csv('StormData.csv')
save( s, file = 'stormDataRaw.Robj' )
```

Exploratory Data Analysis and Cleanup
-------------------------------------

Check for bad data:

```{r check for bad data, cache = TRUE }
for( colName in colnames( s ) ){
  nNA <- sum( is.na( s[,colName] ) )
  if( nNA > 0 ) message( colName, ': ', nNA, ' missing values' )
}
```

None of those are really relevant to the question being asked.

As I have a limited time to do this analysis, I need to work with the minimum
necessary dataset. At the moment it contains `r length( s$EVTYPE )` records.
So I want events with the following __non-zero__

* FATALITIES
* INJURIES
* PROPDMG
* CROPDMG

```{r Filter nonfatal/injurious/damaging events}
wantedEvents <- ( s$FATALITIES>0 | s$INJURIES>0 | s$PROPDMG>0 | s$CROPDMG>0 )
s <- s[wantedEvents,]
```

Now we're down to `r length( s$EVTYPE )` events.

Now to tidy up the damage values. The problem is the multipliers.

```{r silly multipliers}
levels(s$PROPDMGEXP)
levels(s$CROPDMGEXP)
```
