Health and Economic Impact of Severe Weather Events in the USA
==============================================================

__Work by:__ Brian Davies

Executive Summary
-----------------

This report presents an impact analysis for severe weather events in the
Continental US for use by response managers and policymakers.

Following data download, and loading, all events with no health or economic
impact were discarded immediately. After considerable cleanup and exploratory 
data analysis, it was decided to
concentrate on the years from 1993 onwards because the earlier data were thin
and incomplete. This gave a dataset of some 227,000 events. The categorization
was simplified from approximately 440 to approximately 30 event
classes.

Analysing the final clean dataset showed that some 80% of health impacts
(fatalities and injuries) and economic damage (property and
crop damage) were due to only
five types of weather event in each case. However, the annual frequency and
aggregate impact of these event types varied by an order of magnitude from
one year to the next during the period under study.

Software Used
-------------

This work was done using `r R.version.string` with the following libraries loaded:

```{r load necessary library}
library('R.utils')
library('xtable')
```

Getting and loading the online data (cached)
--------------------------------------------

Data were accessed from the sources mentioned in the study brief.

```{r Get and Load Online Data, cache = TRUE }
qtDload = TRUE
download.file( 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2',
               destfile = 'StormData.csv.bz2', method='curl', quiet = qtDload )
download.file( 'https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf',
               destfile = 'pd01016005curr.pdf', method='curl', quiet = qtDload )
download.file( 'https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf',
               destfile = 'Storm_Events_FAQ.pdf', method='curl', quiet = qtDload )
# install.packages("R.utils")
library('R.utils')
bunzip2( 'StormData.csv.bz2', remove = FALSE, overwrite = TRUE )
s <- read.csv('StormData.csv')
save( s, file = 'stormDataRaw.Robj' )
```

Rough Cleanup
-------------

First we check for bad data:

```{r check for bad data, cache = TRUE }
for( colName in colnames( s ) ){
  nNA <- sum( is.na( s[,colName] ) )
  if( nNA > 0 ) message( colName, ': ', nNA, ' missing values' )
}
```

None of those are really relevant to the question being asked.

As I have a limited time to do this analysis, I need to work with the minimum
necessary dataset. At the moment it contains `r length( s$EVTYPE )` records.
So I want events with the following __non-zero__

* FATALITIES
* INJURIES
* PROPDMG
* CROPDMG

```{r Filter nonfatal/injurious/damaging events, cache=TRUE}
wantedEvents <- ( s$FATALITIES>0 | s$INJURIES>0 | s$PROPDMG>0 | s$CROPDMG>0 )
s <- s[wantedEvents,]
```

Now we're down to `r length( s$EVTYPE )` events.

Clean up damage values
----------------------

Now to tidy up the damage values. The problem is the multipliers.

```{r silly multipliers}
table(toupper(as.character(s$PROPDMGEXP))[s$PROPDMG!=0])
table(toupper(as.character(s$CROPDMGEXP))[s$CROPDMG!=0])
```

Will extract these to an array and work there. Set up list of valid __*DMGEXP*__
values first.

```{r fix damage value to numerics, cache=TRUE}
goodexp <- c( 'H', 'K', 'M', 'B' )

p_mantissa <- s$PROPDMG
p_exp <- toupper(as.character(s$PROPDMGEXP))
p_mult <- rep(1,length(p_exp))
p_mult[p_exp=='H'] <- 100
p_mult[p_exp=='K'] <- 1000
p_mult[p_exp=='M'] <- 1e6
p_mult[p_exp=='B'] <- 1e9
p_fixed <- p_mantissa * p_mult
p_fixed[ !(p_exp %in% goodexp) & p_fixed > 0 ] <- NA
summary(p_fixed)

c_mantissa <- s$CROPDMG
c_exp <- toupper(as.character(s$CROPDMGEXP))
c_mult <- rep(1,length(c_exp))
c_mult[c_exp=='H'] <- 100
c_mult[c_exp=='K'] <- 1000
c_mult[c_exp=='M'] <- 1e6
c_mult[c_exp=='B'] <- 1e9
c_fixed <- c_mantissa * c_mult
c_fixed[ !(c_exp %in% goodexp) & c_fixed > 0 ] <- NA
summary(c_fixed)
```

Some of the values look a bit fishy. Specifically, was there REALLY a single
event that cost dollar `r max(p_fixed,na.rm=TRUE)` in property
damage when the all-time total for property damage was
dollar `r sum(p_fixed,na.rm=TRUE)` ?

```{r output fishy hi-cost event}
s[which.max(p_fixed),]
```

A search in Wikipedia for __Napa River Flood__ found a flood in 1986 at
http://en.wikipedia.org/wiki/Napa_River_flood_of_1986 accessed 14 July 2014
which includes the text _"damages totaled $100 million ... Another flood of
lesser proportions occurred on December 31, 2005 after over a week of rain"_
which must mean that the event in the database has got a _B_ instead of an
_M_ in the PROPDMGEXP column. So I'm going to edit out JUST that specific
one (because it's almost a quarter of the total property damage so it's really
egregious), but I don't have the capacity to fix any others.

```{r remove single outlier event}
p_fixed[ !is.na( p_fixed )
         & p_fixed > 1e11 ] <-
  p_fixed[ !is.na( p_fixed)
           & p_fixed > 1e11 ] / 1000
```

Now let's look at the next outlier

```{r output next highest hi-cost event, cache = TRUE}
s[ which.max( p_fixed ), ]
```

That's Katrina, credible at  dollars 31.3 billion
http://en.wikipedia.org/wiki/Hurricane_katrina#Economic_effects accessed
14 July 2014.

What about top crop damage?

```{r Look at max crop damage, cache = TRUE }
max( c_fixed, na.rm = TRUE )
substr(s[which.max(c_fixed),]$REMARKS, 1, 300 )
```

Well, that definitely happened. http://en.wikipedia.org/wiki/Great_flood_of_93#Costs_and_damage
accessed 14 July 2014.

__That's all the time I have available to spend on fixing up damage estimates__

Normalize event categories
--------------------------

Set up a smaller frame __c__ to work with

```{r set up frame c for simplification, cache = TRUE }
evType <- sub( '^ +', '', toupper(as.character(s$EVTYPE)))
c <- data.frame(
  EVTYPE_OLD = evType,
  EVTYPE_NEW = rep( 'other', length(evType)),
  FATALITIES = s$FATALITIES,
  INJURIES = s$INJURIES,
  PROPDMG = p_fixed,
  CROPDMG = c_fixed
  )
c$EVTYPE_OLD <- as.character(c$EVTYPE_OLD)
c$EVTYPE_NEW <- as.character(c$EVTYPE_NEW)
head(c)
```

Add some functions to help monitor progress

```{r some helper functions for cleaning up categories}
# SumCat -- summarize impact of all events matching pattern
SumCat <- function( pattern, set = 'OLD' ){
if( pattern != 'other' ) pattern <- toupper(pattern)
set <- toupper(set)
which <- grep( pattern, c$EVTYPE_OLD )
ntype <- length( levels( as.factor( c$EVTYPE_OLD[which] ) ) )
if( set == 'NEW' ){
  which <- grep( pattern, c$EVTYPE_NEW )
  ntype <- length( levels( as.factor( c$EVTYPE_NEW[which] ) ) )
  }
message( 'In ', set, ' pattern ', pattern, ' matches: ', length( which ), ' events, of ', ntype, ' EVTYPEs' )
message( '   ', sum( c$FATALITIES[which] ), ' FATALITIES' )
message( '   ', sum( c$INJURIES[which] ), ' INJURIES' )
message( '   ', sprintf( '%.2e', sum( c$PROPDMG[which], na.rm=TRUE ) ), ' PROPDMG' )
message( '   ', sprintf( '%.2e', sum( c$CROPDMG[which], na.rm=TRUE ) ), ' CROPDMG' )
}

# Matches() -- enumerate event type matching pattern
Matches <- function( pattern ){
levels( as.factor( evType[ grepl( toupper(pattern), evType ) &
                          ( c$EVTYPE_NEW == 'other' )
                          ] 
                   ) 
        )
}

# Nmatch() -- how many
Nmatch  <- function( pattern ){ length( Matches( pattern ) ) }

# Others() -- which not done yet
Others  <- function(){ levels( as.factor( evType[ c$EVTYPE_NEW == 'other' ] ) ) }

# SetNew() -- allocate new name
SetNew <- function( theFrame, patternFrom, stringTo ){
patternFrom = toupper( patternFrom )
stringTo <- toupper( stringTo )
mlist <- grepl( patternFrom, theFrame$EVTYPE_OLD ) &
  ( theFrame$EVTYPE_NEW == 'other' )
message( 'Replacing ', sum(mlist), ' values' )
theFrame$EVTYPE_NEW[mlist] <- stringTo
theFrame
}
```

So now we do stuff like this

```{r simplify tornadoes TORN, cache = TRUE }
Matches('TORN')
SumCat('TORNADO')
c <- SetNew( c, 'TORN', 'TORNADO' )
```

```{r simplify floods FLOOD, cache = TRUE }
Matches('FLOOD')
SumCat('FLOOD')
c <- SetNew( c, 'FLOOD', 'FLOOD' )
```

```{r simplify hurricanes HURRI, cache = TRUE }
Matches('HURRI')
SumCat('HURRI')
c <- SetNew( c, 'HURRI', 'HURRICANE' )
```

```{r simplify tropical storms and depressions, cache = TRUE }
Matches('TROPI')
SumCat('TROPI')
c <- SetNew( c, 'TROPI', 'TRPCL STORM' )
```

Can't just replace __TSTM__ with __TRPCL STORM__, alas ...

```{r some tropical storms and others TSTM, cache = TRUE }
Matches('TSTM')
c <- SetNew( c, '^NON.TSTM', 'WIND' )
SumCat('TSTM')
c <- SetNew( c, 'TSTM', 'TRPCL STORM' )
```

```{r fixing SNOW, cache = TRUE }
Matches('SNOW')
SumCat('SNOW')
c <- SetNew( c, 'SNOW', 'SNOW' )
```

```{r fixing THUNDER, cache = TRUE }
Matches('THUNDER')
SumCat('THUNDER')
# Someone spells badley
c <- SetNew( c, 'THUNDER',   'THUNDERSTORM' )
c <- SetNew( c, 'THUNDEER',  'THUNDERSTORM' )
c <- SetNew( c, 'THUDER',    'THUNDERSTORM' )
c <- SetNew( c, 'THUNER',    'THUNDERSTORM' )
c <- SetNew( c, 'TUNDER',    'THUNDERSTORM' )
```

__HEAT__ seems to kill a lot of people ...

```{r fixing HEAT, cache = TRUE }
Matches('HEAT')
SumCat('HEAT')
c <- SetNew( c, 'HEAT', 'HEATWAVE' )
```

```{r fixing CHILL, cache = TRUE}
Matches('CHILL')
SumCat('CHILL')
c <- SetNew( c, 'CHILL', 'WINDCHILL' )
```

```{r fixing HAIL, cache = TRUE}
Matches('HAIL')
SumCat('HAIL')
c <- SetNew( c, 'HAIL', 'HAIL' )
```

Can't believe I missed this one ...

```{r fixing WIND, cache = TRUE}
Matches('WIND')
SumCat('WIND')
c <- SetNew( c, 'WIND', 'WIND' )
```

```{r fixing LIGHTNING, cache = TRUE}
Matches('LIG')
SumCat('LIG')
c <- SetNew( c, 'LIG', 'LIGHTNING' )
```

```{r fixing wintry conditions in general, cache = TRUE}
Matches('WINT')
SumCat('WINT')
c <- SetNew( c, 'WINT', 'WINTRY OTHER' )
```

```{r fixing fires in general, cache = TRUE}
Matches('FIRE')
SumCat('FIRE')
c <- SetNew( c, 'FIRE', 'FIRES' )
```

Following command gives top n remaining in category

```{r filter highest-scoring event in category }
WorstN <- function( category, n = 10 ){
o <- c[c$EVTYPE_NEW=='other', ];
o <- o[ order( o[, toupper( category ) ], decreasing=TRUE ), ];
head( o, n )
}
WorstN( 'injuries', 12 )
```

Consolidate a few more ...

```{r some more categories, cache = TRUE }
c <- SetNew( c, 'BLIZZ', 'BLIZZARD' )
c <- SetNew( c, 'DROUGHT', 'DROUGHT' )
c <- SetNew( c, 'ICE STORM', 'ICE STORM' )
c <- SetNew( c, 'STORM SURGE', 'STORM SURGE' )
c <- SetNew( c, 'UNSEASONABLY WARM', 'HEATWAVE' )
c <- SetNew( c, 'COLD', 'XTRM COLD' )
c <- SetNew( c, 'LOW TEMPERATURE', 'XTRM COLD' )
c <- SetNew( c, 'FOG', 'FOG' )
c <- SetNew( c, 'RIP', 'RIP CURRENT' )
c <- SetNew( c, 'DUST', 'DUST STORM' )
c <- SetNew( c, 'SURF', 'SURF' )
c <- SetNew( c, 'LANDSLIDE', 'LANDSLIDE' )
c <- SetNew( c, 'MUDSLIDE', 'MUDSLIDE' )
c <- SetNew( c, 'AVALANCH?E', 'AVALANCHE' )
c <- SetNew( c, 'HYPOTHERMIA', 'HYPOTHERMIA' )
```

Non-freezing rain ...

```{r Most rain,  cache = TRUE }
Matches('RAIN')
Matches('^[^F].*RAIN')
SumCat('^[^F].*RAIN')
c <- SetNew( c, '^[^F].*RAIN', 'RAIN' )
```
Some ice phenomena ...

```{r unpick & fix actual ice, cache = TRUE }
WorstN('injuries')
Matches('ICE|GLAZE')
c <- SetNew( c, 'GLAZE', 'ICE' )
c <- SetNew( c, 'ICE.+ROAD', 'ICE' )
c <- SetNew( c, '^ICE$', 'ICE' )
c <- SetNew( c, 'BLACK ICE', 'ICE' )
```

Some tidying up __FLD__ = 'FLOOD' ...

```{r Fix FLD, cache = TRUE }
Matches('FLD')
c <- SetNew( c, 'FLD', 'FLOOD' )
```

Maritime grab-bag ...

```{r fix offshore, cache = TRUE }
c <- SetNew( c, 'SEA', 'MARITIME' )
c <- SetNew( c, 'MARINE MISHAP', 'MARITIME' )
c <- SetNew( c, 'MARINE ACCIDENT', 'MARITIME' )
c <- SetNew( c, 'TSUNAMI', 'TSUNAMI' )
c <- SetNew( c, 'WATERSPOUT', 'WATERSPOUT' )
```

Finished classifying events!

```{r Final summary of classifications}
# Everything not otherwise classified ...
SumCat( 'other', 'new' )
# List of new classes
levels(as.factor(c$EVTYPE_NEW))
```

That's `r length( levels(as.factor(c$EVTYPE_NEW)))` categories, down from
`r length(levels(as.factor(evType)))` in the filterd dataset.

Get a nice year column
----------------------

Just want the 4-digit year.

```{r get vector of integer years, cache = TRUE}
dateOrig <- as.character(s$BGN_DATE)
nDates <- length( dateOrig )
dateTok <- strsplit( dateOrig, '/| ' )
years <- rep( NA, nDates )
for( i in seq( 1, nDates ) ){ years[i] <- as.integer( dateTok[[i]][3] ) }
summary( years )
c$YEAR <- years
head( c )
```

There's something a bit fishy about the event frequencies through time.
Here goes a nice summary table ...

```{r summary of data by year, cache = TRUE }
notypes <- data.frame( FATALITIES = c$FATALITIES, INJURIES = c$INJURIES,
                       PROPDMG = c$PROPDMG, CROPDMG = c$CROPDMG )

yInc <- aggregate( c$YEAR, list(c$YEAR), length )
yTot <- aggregate(notypes,list(c$YEAR), sum, na.rm=TRUE )
ySum <- data.frame(
  YEAR = yInc$Group.1,
  N = yInc$x,
  DEATHS = yTot$FATALITIES,
  INJURIES = yTot$INJURIES,
  PROPDMG = yTot$PROPDMG,
  CROPDMG = yTot$CROPDMG
  )
ySum
```

Everything changed in 1993. Deaths jumped, and __CROPDMG__ started.

__So I'm going to ignore everything before 1993__

```{r edit out everything pre-1993}
c <- c[c$YEAR>=1993,]
```

Now for a summary of stats by event type

```{r summary of post-1992 by type, cache = TRUE }
notypes <- data.frame( FATALITIES = c$FATALITIES, INJURIES = c$INJURIES,
                       PROPDMG = c$PROPDMG, CROPDMG = c$CROPDMG )

evCnt <- aggregate( c$EVTYPE_NEW, list(c$EVTYPE_NEW), length )
evTot <- aggregate(notypes,list(c$EVTYPE_NEW), sum, na.rm=TRUE )
evSum <- data.frame(
  EVTYPE = evCnt$Group.1,
  N = evCnt$x,
  FATALITIES = evTot$FATALITIES,
  INJURIES = evTot$INJURIES,
  PROPDMG = evTot$PROPDMG,
  CROPDMG = evTot$CROPDMG
  )
evSum
```

Analysis of Health Impacts
--------------------------

Now I want to summarize the n worst CATEGORIES since 1993 for __FATALITIES__
and __INJURIES__. The best way to do this is by ranking the different event
types by _descending_ impact (number of DEATHS or FATALITIES) and choosing
the top few. As a nod to quality we'll monitor what percent of the total we
are capturing in these top few.

```{r choose how many categories}
nCat <- 5
```

After some trial and error I went for `r nCat` categories in each case.

```{r define WorstCat }
WorstCat <- function( theFrame, impact, n = 5, format = NULL ){
  subFrame <- theFrame[, c('EVTYPE','N',impact)]
  catOrd <- order( theFrame[,impact], decreasing = TRUE )[1:n]
  nWorstFrame <- subFrame[catOrd,]
  
  nSumFrame <- data.frame(
    EVTYPE = c(''),
    N = c( sum( nWorstFrame$N)),
    IMPACT = c(sum( nWorstFrame[, impact]))    
  )
  colnames(nSumFrame) <- c( 'EVTYPE', 'N', impact )
  
  totalsFrame <- data.frame(
    EVTYPE = c(''),
    N = c( sum( subFrame$N)),
    IMPACT = c(sum( subFrame[, impact]))
  )
  colnames(totalsFrame) <- c( 'EVTYPE', 'N', impact )
  
  othersFrame <- data.frame(
    EVTYPE = c(''),
    N = c( totalsFrame[1,'N'] - sum( nWorstFrame[,'N']) ),
    IMPACT = c( totalsFrame[1,impact] - sum( nWorstFrame[,impact]) )
  )
  colnames(othersFrame) <- c( 'EVTYPE', 'N', impact )
  
  rankFrame <- data.frame(
    RANK = as.character( c( c(1:n),
                            paste0('Top ', n),
                            'Other',
                            'TOTAL' 
    )
    )
  )
  
  reportFrame <- rbind(nWorstFrame, nSumFrame, othersFrame, totalsFrame )
  
  pcTotFrame <- data.frame(
    FOO <- reportFrame[,impact]/totalsFrame[1,impact]*100
  )
  colnames(pcTotFrame) <- c( paste0( '%', impact ) )
  
  reportFrame <- cbind( rankFrame, reportFrame, pcTotFrame )
  
  if( ! is.null( format )){
    reportFrame[,impact] <- sprintf( format, reportFrame[,impact ] )
  }
  
  reportFrame
}
```

```{r Find biggest causes of fatalities, results = 'asis' }
worstFatalities <- WorstCat( evSum, 'FATALITIES', 5 )
myxt <- xtable(worstFatalities,
        digits = c( 0, 0, 0, 0, 0, 1 ),
        caption = 'Causes of weather-related FATALITIES post-1992'
        )
print( myxt, type = 'html', include.rownames = FALSE )
```

```{r Find biggest causes of injuries, results = 'asis' }
worstInjuries <- WorstCat( evSum, 'INJURIES', 5 )
myxt <- xtable(worstInjuries,
        digits = c( 0, 0, 0, 0, 0, 1 ),
        caption = 'Causes of weather-related INJURIES post-1992'
        )
print( myxt, type = 'html', include.rownames = FALSE )
```

So the worst event types for __FATALITIES__ are
`r worstFatalities$EVTYPE[1:nCat]`
, accounting for
`r worstFatalities[nCat+1,'%FATALITIES']`
percent, and the worst for __INJURIES__ are
`r worstInjuries$EVTYPE[1:5]`
, accounting for 
`r worstInjuries[nCat+1,'%INJURIES']`
percent of the total.

We can do a multipanel graph showing the time evolution of these accident types.

```{r Plot health impacts, warning = FALSE }
onePlot <- function(
  sourceFrame = NULL,
  aggFunc = 'sum',
  impact = 'FATALITIES',
  evTypes = NULL,
  verbose = FALSE
){
  colors <- c( 'black', 'red', 'green', 'blue', 'orange' )
  Years <- levels(as.factor(sourceFrame$YEAR))
  plotFrame <- data.frame( YEAR = Years )
  # Assemble data frame for plotting so we can yet scale
  nEvt <- length( evTypes )
  for( iEv in (1:nEvt) ){
    eVtype <- evTypes[iEv]
    if( verbose) message( 'Doing ', eVtype )
    typeFrame <- sourceFrame[sourceFrame$EVTYPE_NEW==eVtype,]
    eVFrame <- data.frame(
      YEAR = typeFrame$YEAR,
      X = typeFrame[,impact]
    )
    colnames(eVFrame) <- c( 'YEAR', impact )
    if( aggFunc == 'sum' ) sumFrame <- aggregate(eVFrame, list(eVFrame$YEAR), sum, na.rm=TRUE)
    if( aggFunc == 'length' ) sumFrame <- aggregate(eVFrame, list(eVFrame$YEAR), length)
    colnames(sumFrame) <- c( 'YEAR', 'foo', eVtype )
    sumFrame$YEAR <- as.factor(sumFrame$YEAR)
    plotFrame <- merge( plotFrame, sumFrame[,c('YEAR',eVtype)], by = 'YEAR', all.x = TRUE )
    plotFrame[which(is.na(plotFrame[,eVtype])),eVtype] = 0
  } 
  yMax <- max(plotFrame[,2:(nEvt+1)])
  if( aggFunc == 'sum' ){
    mainTitle <- paste0( "Annual ", impact, " -- Major Event types" )
    yLab <- paste0( "Total ", impact, " in year")
  }
  if( aggFunc == 'length' ){
    mainTitle <- paste0( "Annual events with ", impact )
    yLab <- paste0( "Number in Year" )
  }
  plot(plotFrame$YEAR, plotFrame[,2], type='n', lwd=2, col=colors[iEv],
     xlab = 'YEAR', ylim = c( 0, yMax ), ylab = yLab, main = mainTitle, lwd = 0 )
  for( iEv in (1:nEvt) ){
    points(plotFrame$YEAR, plotFrame[,iEv+1], type='l', lwd=2, col=colors[iEv] ) 
  }
  legend( x = 'topright', evTypes, lwd = 2, col=colors )
}

HealthPlot <- function ( theFrame, eVsets = NULL ){
  
  par(mfrow=c(2,2))
  
  onePlot(sourceFrame = theFrame, aggFunc = 'sum', impact = 'FATALITIES',
          evTypes = eVsets[[1]])
  onePlot(sourceFrame = theFrame, aggFunc = 'sum', impact = 'INJURIES',
          evTypes = eVsets[[2]])
  onePlot(sourceFrame = theFrame, aggFunc = 'length', impact = 'FATALITIES',
          evTypes = eVsets[[1]])
  onePlot(sourceFrame = theFrame, aggFunc = 'length', impact = 'INJURIES',
          evTypes = eVsets[[2]])
}

HealthPlot( c, eVsets = list( c( as.character(worstFatalities$EVTYPE[1:nCat]) ),
                              c( as.character(worstInjuries$EVTYPE[1:nCat]) ) ) )
```

Analysis of Economic Impacts
----------------------------

Similar approach to before, worst `r nCat` categories.

```{r Find biggest causes of property damage, results = 'asis' }
worstPropDmg <- WorstCat( evSum, 'PROPDMG', 5, format = '%.3e' )
myxt <- xtable(worstPropDmg,
        digits = c( 0, 0, 0, 0, 0, 1 ),
        caption = 'Causes of weather-related Property Damage post-1992'
        )
print( myxt, type = 'html', include.rownames = FALSE )
```

```{r Find biggest causes of crop damage, results = 'asis' }
worstCropDmg <- WorstCat( evSum, 'CROPDMG', 5 )
myxt <- xtable(worstCropDmg,
        digits = c( 0, 0, 0, 0, 0, 1 ),
        caption = 'Causes of weather-related Crop Damage post-1992'
        )
print( myxt, type = 'html', include.rownames = FALSE )
```

So the worst event types for __PROPDMG__ are
`r worstPropDmg$EVTYPE[1:nCat]`
, accounting for
`r worstPropDmg[nCat+1,'%PROPDMG']`
percent, and the worst for __CROPDMG__ are
`r worstCropDmg$EVTYPE[1:5]`
, accounting for 
`r worstCropDmg[nCat+1,'%INJURIES']`
percent of the total.

We can do a multipanel graph showing the time evolution of these accident types.

```{r Plot economic impacts, warning = FALSE }
HealthPlot( c, eVsets = list( c( as.character(worstPropDmg$EVTYPE[1:nCat]) ),
                              c( as.character(worstCropDmg$EVTYPE[1:nCat]) ) ) )
```

Conclusions from Statistical Analysis
-------------------------------------